A novel approach to chemical tagging in lower-dimensional latent space

### 2018-12-18 article/ms.tex
introduction + citations

Chemical tagging seeks to identify star formation events using the present
day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
The detailed chemical abundances observable in a star's photosphere are a
fossil record that carries with it information about where and when a star
formed. While the photospheric abundances remain largely unchanged throughout
a star's lifetime \citep[however see][]{}, the dynamical dissipation timescale
of open clusters is only about 1~Gyr \citep{timescale}. That makes chemical
tagging an attractive approach to identify star formation sites long after
those stars are no longer gravitationally bound.

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Early works 
works examined metrics quantifying the similarity of chemical abundances
in gravitationally bound star clusters, and more

 at the similarity of chemical abundance metrics 
Theoretical works have looked at the dimensionality

From
theoretical studies examining the

 the dynamical dissipation timescale of open clusters is only
about 1~Gyr \citep{timescale} and the photospheric abundances remain largely
unchanged throughout a star's lifetime \citep[however see][]{}, the ability
to identify star formation sites long after those stars are no longer
gravitationally bound.

The \Galah\ survey \citep{DaSilva:2015} is a stellar spectroscopic survey 
that uses the High Efficiency and Resolution Multi-Element 
Spectrograph \citep[HERMES][]{Sheinis:2016} on the Australian Astronomical Telescope (AAT).
The survey will observe up to $10^6$ stars in the Milky Way, and
measure up to 30 chemical abundances for each star. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). These data provide an unparalleled view on the production
of chemical elements in the Milky Way.

\todo{early approaches on clustering metrics}

\todo{work on lower limits of chemical homogeneity}

All clustering techniques applied to chemical tagging assume that the data
dimensions are independent. However, theory and observations agree this cannot
be true. Nucleosynthetic processes produce multiple elements in varying
quantities, and the dimensionality of stellar abundance datasets has been shown
to be lower than the actual number of dimensions. Any clustering approach that
treats each new elemental abundance as an indenpendent axis of information will
therefore conclude with biased inferences about the star formation history of
our Galaxy. On the other hand, it is difficult to prescribe the nucleosynthetic
yields that have contributed to the chemical abundances of each star. There are
qualitative statements that can be made for large numbers of stars, or particular
types of stars, but quantifying the precise contribution of different processes
to each star is an unsolved problem. For example, the [$\alpha$/Fe] `knee' in
abundance ratios in the Milky Way can qualitatively be explained by 
core-collapse supernovae being the predominant nucleosynthetic process in the
early Milky Way before Type Ia contributions had a significant impact, but 
efforts to date have not sought to try to explain the detailed abundances of a 
single star as a contribution of yields from different systems (e.g., 20\% of the
Ba was produced by 3 M$_\odot$ systems). This is in part because of the
challenging and degenerate nature of the problem as described, and is complicated
by the differences in yield predictions that account from prescriptions used in
different codes.

In this work we present a new approach to chemical tagging that allows us to 
identify the latent factors that contribute to the chemical abundances of all 
stars (e.g., something akin to nucleosynthetic yields), while simultaneously
performing clustering in the latent space. Notwithstanding caveats that we will
discuss in detail, this allows us to infer nucleosynthetic yields rather than
strictly prescribe them from models. Moreover, the scale of the clustering
problem reduces by a significant fraction because the clustering is performed in
a lower dimensional latent space instead of the higher dimensional data space.
In Section~\ref{sec:methods} we describe the model and the methods we use to
estimate the model parameters. Section~\ref{sec:experiments} describe the 
experiments performed using generated and real data sets. We discuss the results
of these experiments in Section~\ref{sec:discussion}, including the caveats with
the model as described. We conclude in Section~\ref{sec:conclusion}

@@

Factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep{factor-analysis}.
Related techniques include principal component analysis \citep{pca} and its
variants \citep{ppca}, singular value decomposition \citep{svd}, and other
matrix factorization methods. While factor analysis on its own is a useful
dimensionality reduction tool to identify latent factors that contribute to
the chemical abundances of stars, it does not perform any clustering.
Similarly, clustering techniques applied to chemical abundances to date 
\citep[e.g.,][]{Hogg:2016} do not account for the lower dimensionality of the
data.

Here we use a combination of factor analysis and a gaussian mixture model to
provide a consistent description of the data, given the latent variables and 
their clustering in latent space. In this work the data $\vecdata$ is a 
$\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
stars and $\NumDimensions$ is the number of chemical abundances measured 
for each star. We assume a generative model for the data 
\begin{equation}
	\vecdata = \factorloads\factorscores + \vec{e}
	\label{eq:generative-model}
\end{equation}

@@

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, and the
Bayesian Information Criterion \citep{bic}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad . \label{eq:bic}
\end{equation}

@@

In this experiment we use detailed chemical abundances of metal-poor stars as
reported by \citet{Barklem:2003}, and we demonstrate the utility and limitations
in interpreting the latent factors as nucleosynthetic processes. The data 
contains 14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak),
as well as Sr, Y, Ba, and Eu ($s$/$r$-process). We performed a grid search for
the optimal number of latent factors $J$ and components $K$, starting from
$J = 1$ and $K = 1$, and trialling up until $J = 8$ and $K = 5$. We initialised
the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
and BIC recorded for the optimised set of model parameters for each permutation 
of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
five latent factors are preferred, and no clustering.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2003} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
			 The cross at
			 $J=8$ and $K=5$ indicates that the model could not converge without
			 adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $J = 5$ and $K = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2003}
			 data. \todo{elaborate latent factors}}
    \label{fig:exp2-latent-factors}
\end{figure*}

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-specific-scatter.png}
	\caption{Specific scatter remaining in the \citet{Barklem:2003} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}

>>>>

Chemical tagging seeks to identify star formation events using the present
day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2002}.
The detailed chemical abundances observable in a star's photosphere provide a
fossil record that carries with it information about where and when a star
formed. While the photospheric abundances remain largely unchanged throughout
a star's lifetime \citep[however see][]{Dotter:2017,Ness:2018b}, the dynamical 
dissipation timescale of open clusters in the Milky Way disc is of order a few 
gigayears \citep{Portegies-Zwart:1998}. That makes chemical tagging an attractive 
approach to identify star formation sites long after those stars are no longer 
gravitationally bound.

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Although biases arise when
only considering star clusters that are still gravitationally bound, the chemical
homogenity of open clusters provides an empirical measure of how similar stars
would need to be before they could be tagged as belonging to the same
star formation site \citep{Mitschang:2014}. However, there are also analysis
issues in understanding how well those chemical abundances can be measured
\citep{Bovy:2016}. If open clusters were truely chemically homogeneous then our
ability to chemically tag the Milky Way depends on the precision with which we
can measure those chemical abundances in stars. Data-driven techniques are
driving a revolution in this field \citep{Ness:2015,Ness:2018a,Ness:2018b,
Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but there is more work
to do: astronomers have not yet reached the Cram\'er-Rao bound for chemical 
abundances of stars.

There are further chemical tagging obstacles to overcome once all the chemical 
abundance information in a stellar spectrum can be extracted. Most notably is a 
catalogue of precise detailed chemical abundances for a large number of stars,
where those chemical abundances trace different nucleosynthetic pathways.
This is the primary scientific goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
a stellar spectroscopic survey that uses the High Efficiency and Resolution 
Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the Australian 
Astronomical Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
Milky Way, and measure up to 30 chemical abundances for each star. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). These data provide an unparalleled view on the production
of chemical elements in the Milky Way.

Given these data and the most
favourable assumptions in chemical tagging -- that star clusters are truely
chemically homogenous, and we can measure those abundances with infinite precision,
and those abundances are differentiable between star clusters -- then chemical
tagging becomes a clustering problem. All clustering techniques applied to 
chemical tagging assume that the data dimensions are independent. That is to say
that adding a dimension of [Ni/H] provides totally new, independent information.
Theory and observations agree that this cannot be true.
Nucleosynthetic processes produce multiple elements in varying
quantities, and the dimensionality of stellar abundance datasets has been shown
to be lower than the actual number of dimensions \citep{Ting:2012, Price-Jones:2018}.
Any clustering approach that treats each new elemental abundance as an 
indenpendent axis of information will therefore conclude with biased inferences
about the star formation history of our Galaxy.

Alternative approaches are perhaps more challenging, as it is difficult to prescribe
the nucleosynthetic yields that have contributed to the chemical abundances of each star
with great confidence. There are
qualitative statements that can be made for large numbers of stars, or particular
types of stars, but quantifying the precise contribution of different processes
to each star is an unsolved problem. For example, the [$\alpha$/Fe] `knee' in
abundance ratios in the Milky Way can qualitatively be explained by 
core-collapse supernovae being the predominant nucleosynthetic process in the
early Milky Way before Type Ia contributions had a significant impact, but 
efforts to date have not sought to try to explain the detailed abundances of a 
single star as a contribution of yields from different systems (e.g., 20\% of the
Ba was produced by 3 M$_\odot$ systems). This is in part because of the
challenging and degenerate nature of the problem as described, and is complicated
by the differences in yield predictions that account from prescriptions used in
different codes.

New approaches to chemical tagging are clearly needed. Immediate advances would
include methods that take the dependence among chemical elements into account
within some generative model, or techniques that combine chemical abundances
with dynamical constraints to place joint prior probabilities on whether any
two stars could have formed from the same star cluster, given some model of the
Milky Way. In this work we focus on the former. Here we present a new approach
to chemical tagging that allows us to identify the latent (unobserved) factors
that contribute to the chemical abundances of all stars (e.g., nucleosythetic
yields) while simultaneously performing clustering in the latent space.
Notwithstanding caveats that we will
discuss in detail, this allows us to infer nucleosynthetic yields rather than
strictly prescribe them from models. Moreover, the scale of the clustering
problem reduces by a significant fraction because the clustering is performed in
a lower dimensional latent space instead of the higher dimensional data space.
In Section~\ref{sec:methods} we describe the model and the methods we use to
estimate the model parameters. Section~\ref{sec:experiments} describe the 
experiments performed using generated and real data sets. We discuss the results
of these experiments in Section~\ref{sec:discussion}, including the caveats with
the model as described. We conclude in Section~\ref{sec:conclusion}.

@@

Factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
Related techniques include principal component analysis \citep{Hotelling:1933} and its
variants \citep{Tipping;Bishop:1999}, singular value decomposition \citep{Golub:1970}, and other
matrix factorization methods. While factor analysis on its own is a useful
dimensionality reduction tool to identify latent factors that contribute to
the chemical abundances of stars \citep[e.g.,][]{Price-Jones:2018}, it does not perform any clustering.
Similarly, clustering techniques applied to chemical abundances to date 
\citep[e.g.,][]{Hogg:2016} do not account for the lower dimensionality of the
data.

Here we use a variant of factor analysis known elsewhere as a mixture of common 
factor analyzers \citep{Baek:2010}, where the data are generated by a set of 
latent factors that are common to all data, but the scoring (or extent) of those
factors is different for each data point, and the data can be modelled as a
mixture of multivariate normal distributions in the latent space (factor scores).
In this work the data $\vecdata$ is a 
$\NumData \times \NumDimensions$ matrix where $\NumData$ is the number of 
stars and $\NumDimensions$ is the number of chemical abundances measured 
for each star. We assume a generative model for the data 
\begin{equation}
	\vecdata = \factorloads\factorscores + \vec{e}
	\label{eq:generative-model}
\end{equation}

@@

Here we treat the generated data set as if the true number of latent factors
and the true number of components are not known. Starting with $\NumLatentFactors = 1$
and $\NumComponents = 1$, we trialled each permutation of $\NumLatentFactors$ and $\NumComponents$
until $\NumLatentFactors_{max} = 10$
and   $\NumComponents_{max} = 40$ (e.g., twice $\NumLatentFactors_{true}$ and $\NumComponents_{true}$).
For each $\NumLatentFactors,\NumComponents$ permutation we initialised the factor
loads and component assignments randomly. We performed expectation-maximization 
cycles until the relative log-likelihood improved by less than $10^{-5}$. We then
recorded the log-likelihood of the data given the model parameters, and the
Bayesian Information Criterion \citep{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad . \label{eq:bic}
\end{equation}

@@

In this experiment we use detailed chemical abundances of metal-poor stars as
reported by \citet{Barklem:2005}, and we demonstrate the utility and limitations
in interpreting the latent factors as nucleosynthetic processes. The data 
contains 14 measurements of chemical abundances for 61 stars ($N = 61$, $D = 14$): 
Al (light odd-Z), Mg and Ca ($\alpha$), Sc, Ti, Cr, Mn, Fe, Co, and Ni (Fe-peak),
as well as Sr, Y, Ba, and Eu ($s$/$r$-process). We performed a grid search for
the optimal number of latent factors $J$ and components $K$, starting from
$J = 1$ and $K = 1$, and trialling up until $J = 8$ and $K = 5$. We initialised
the model parameters randomly for each permutation of $J$ and $K$, and ran E-M
until a convergence threshold of $10^{-5}$ was reached in the log likelihood. 
In Figure~\ref{fig:exp2-gridsearch-contours} we show the negative log likelihood 
and BIC recorded for the optimised set of model parameters for each permutation 
of $J$ and $K$. The lowest BIC value is found for $J = 5$ and $K = 1$, indicating
five latent factors are preferred, and no clustering.

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-ll.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-gridsearch-bic.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 \citet{Barklem:2005} data in Experiment~2.
    		 The top panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, and the lower panel shows the BIC (Equation \ref{eq:bic}).
			 The cross at
			 $J=8$ and $K=5$ indicates that the model could not converge without
			 adding a regularisation constant to the diagonals of $\scorecovs$.
			 The marker indicates the lowest value in each panel, showing that
			 $J = 5$ and $K = 1$ is preferred by BIC.}
    \label{fig:exp2-gridsearch-contours}
\end{figure}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-latent-factors.png}
	\caption{Rotated latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. \todo{elaborate latent factors}}
    \label{fig:exp2-latent-factors}
\end{figure*}

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp2-specific-scatter.png}
	\caption{Specific scatter remaining in the \citet{Barklem:2005} data used in
			 Experiment~2 after accounting for the contributions by all latent
			 factors.}
    \label{fig:exp2-specific-scatter}
\end{figure}
